### 2020-03-11 Writing/disparity_review.tex
ref fix
Updated references and affiliations

\bigskip
\noindent RH: disparate disparity analyses
\bigskip
\medskip
\begin{center}
\noindent{\Large \bf Disparities in the analysis of morphological disparity}
\bigskip

Thomas Guillerme$^{1,16,*,+}$,
Natalie Cooper$^{2,+}$,
Stephen L. Brusatte$^{3}$,
Katie E. Davis$^{4}$,
Andrew L Jackson$^{5}$,
Sylvain Gerber$^{6}$,
Anjali Goswami$^{2}$,
Kevin Healy$^{7}$,
Melanie Hopkins$^{8}$,
Marc EH Jones$^{9}$,
Graeme T. Lloyd$^{10}$,
Joseph E. O'Reilly$^{11}$,
Abi Pate$^{11}$,
Mark N Puttick$^{12}$,
Emiliy Rayfield$^{11}$,
Erin E. Saupe$^{13}$,
Emma Sherratt$^{14}$,
Graham Slater$^{15}$,
Vera Weisbecker$^{1}$,
Gavin H Thomas$^{16}$
and Philip Donoghue$^{11}$\\

\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
$^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
$^{4}$Department of Biology, University of York, YO10 5DD, UK;
$^{5}$Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
$^{6}$Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
$^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
$^{8}$ @@@;
$^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
$^{11}$School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
$^{12}$Milner Centre for Evolution, University of Bath, BA2 7AY UK;
$^{13}$Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
$^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
$^{15}$ @@@;
$^{16}$ @@@;\\
$^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding author: guillert@tcd.ie\\}

@@

	\item \textbf{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
	This makes disparity an informative complement to diversity measures based on species richness alone. 
	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
	\citep[e.g.][]{Fortey1996, Ruta2013, Hopkins2013}.
	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011}.

@@

\noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[e.g.][]{Foote1989, Deline2018};
	2) continuous measurements of features (e.g. lengths) \citep[e.g][]{Friedman2010}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[e.g.][]{Cooney2017}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017}.

@@

Disparity analyses often use ordination techniques for dimension reduction.
Ordinations are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g. principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS; etc.).
They come in many flavours depending on the data and the desired morphospace properties.
For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) \citep[see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.

One of the reasons why ordination techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at a time, which can be more intuitive than through disparity indices (see section \ref{section:metrics} below).
Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e. selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012,dryden2016statistical}.
In practice, this allows us to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g. the elongation and flattening of birds beaks;][]{Cooney2017}.

@@

Categorical data are a good deal more problematic, because the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean, and the distribution of the variance is usually more evenly distributed across axes (i.e. contrary to a PCA, the first few axis do not encompass most variance of the dataset).
Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e. the distance between A and B is not equal to the distance between B and A][]{Gerber2014}.
Last, but not least, in many cases, ordination might not be necessary.
For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015}.
For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.

@@

When considering only one dimension, disparity indices can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these indices, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g. volume
\citealt{Diaz2016}, distance from the centroid \citealt{Hopkins2017, Finlay2015-ft}, variance and range \citealt{Brusatte2008}).
Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g. does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) indices related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015} - see section \ref{section:ordination} above).

@@

Furthermore, multidimensional spaces have some counterintuitive properties that should be considered such as the ``curse of dimensionality'' \citep{Bellman1966}.
In spaces with some axes of variance lower than 1, product-based indices used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966}.
Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992} or convex hull based indices \citep{Jackson2011}.

@@

One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in disparity-through-time analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e. the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original data; or (2) post-transformation, i.e. the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g. the ordination scores) \citep{lloyd2018}.

@@

\noindent Morphological disparity analyses are widely employed in evolutionary palaeobiology, and are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse them, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method and ancestral state estimation method on disparity analysis results.
Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003, Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b}.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013, Saupe2015, Canter2018, mammola2019}.
Innovations in morphological disparity analyses likely await discovery in their respective literatures.

@@

Although disparity analyses are now simple to implement in freely available software \citep{Navarro2003, Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b} it is crucial to remember that they are multidimensional analyses and that multidimensional analyses are complex.
We assert that future morphological analyses will benefit from emphasising the methodological decisions made, rather than simply using disparity analysis because it exists.

>>>>

\medskip
\begin{center}
\noindent{\Large \bf Disparities in the analysis of morphological disparity}
\bigskip

Thomas Guillerme$^{1,16,*,+}$,
Natalie Cooper$^{2,+}$,
Stephen L. Brusatte$^{3}$,
Katie E. Davis$^{4}$,
Andrew L Jackson$^{5}$,
Sylvain Gerber$^{6}$,
Anjali Goswami$^{2}$,
Kevin Healy$^{7}$,
Melanie J. Hopkins$^{8}$,
Marc E. H. Jones$^{9}$,
Graeme T. Lloyd$^{10}$,
Joseph E. O'Reilly$^{11}$,
Abi Pate$^{11}$,
Mark N. Puttick$^{12}$,
Emiliy Rayfield$^{11}$,
Erin E. Saupe$^{13}$,
Emma Sherratt$^{14}$,
Graham J. Slater$^{15}$,
Vera Weisbecker$^{1}$,
Gavin H. Thomas$^{16}$
and Philip C. J. Donoghue$^{11}$\\

\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
$^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
$^{4}$Department of Biology, University of York, YO10 5DD, UK;
$^{5}$Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
$^{6}$Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
$^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
$^{8}$ Division of Paleontology, American Museum of Natural History, New York, USA;
$^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
$^{11}$School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
$^{12}$Milner Centre for Evolution, University of Bath, BA2 7AY UK;
$^{13}$Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
$^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
$^{15}$ Department of the Geophysical Sciences, University of Chicago, Chicago, Illinois 60637;
$^{16}$ Department of Animal and Plant Sciences, The University of Sheffield, Sheffield S10 2TN, UK;\\
$^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding author: guillert@tcd.ie\\}

@@

	\item \textbf{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
	This makes disparity an informative complement to diversity measures based on species richness alone. 
	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
	\citep[e.g.][]{Fortey1996, Hopkins2013}.
	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{pierce2008patterns}.

@@

\noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[e.g.][]{Foote1989, Deline2018};
	2) continuous measurements of features \citep[e.g. lengths in][]{Friedman2010}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[e.g.][]{sherratt2017rates}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017}.

@@

Disparity analyses often use ordination techniques for dimension reduction.
Ordinations are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g. principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS; etc.).
They come in many flavours depending on the data and the desired morphospace properties.
For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling \citep[NMDS; see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.

One of the reasons why ordination techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at a time, which can be more intuitive than through disparity indices (see section \ref{section:metrics} below).
Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e. selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012,dryden2016statistical}.
In practice, this allows us to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g. the elongation and flattening of birds beaks;][]{bright2016shapes}.

@@

Categorical data are a good deal more problematic, because the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean, and the distribution of the variance is usually more evenly distributed across axes (i.e. contrary to a PCA, the first few axis do not encompass most variance of the dataset).
Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e. the distance between A and B is not equal to the distance between B and A;][]{Gerber2014}.
Last, but not least, in many cases, ordination might not be necessary.
For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep[e.g.][]{Close2015}.
For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.

@@

When considering only one dimension, disparity indices can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these indices, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g. volume;
\citealt{Diaz2016}, distance from the centroid; \citealt{Hopkins2017, Finlay2015}, variance and range; \citealt{Brusatte2008}).
Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g. does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) indices related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015} - see section \ref{section:ordination} above).

@@

Furthermore, multidimensional spaces have some counterintuitive properties that should be considered such as the ``curse of dimensionality'' \citep{Bellman1966}.
In spaces with some axes of variance lower than 1, product-based indices used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966}.
Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Wills2001} or convex hull based indices \citep{Jackson2011}.

@@

One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in disparity-through-time analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e. the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original data; or (2) post-transformation, i.e. the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix \citep[e.g. the ordination scores;][]{lloyd2018}.

@@

\noindent Morphological disparity analyses are widely employed in evolutionary palaeobiology, and are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse them, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method and ancestral state estimation method on disparity analysis results.
Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b}.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013, Saupe2015, Canter2018, mammola2019}.
Innovations in morphological disparity analyses likely await discovery in their respective literatures.

@@

Although disparity analyses are now simple to implement in freely available software \citep{Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b} it is crucial to remember that they are multidimensional analyses and that multidimensional analyses are complex.
We assert that future morphological analyses will benefit from emphasising the methodological decisions made, rather than simply using disparity analysis because it exists.
